% ai-phishing-detection-dissertation/report/preamble/acknowledgements.tex

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

\noindent I would first and foremost like to express my deepest gratitude to my family. To my wonderful parents, \uline{\textbf{Mum}} and \uline{\textbf{Dad}}, thank you for your unwavering support, encouragement, and belief in me throughout my academic journey and during this challenging project. To my sister, \uline{\textbf{Preethika}}, thank you for your constant support and for always being there.\newline

\noindent I would also like to extend my sincere thanks to my friends and peers who provided support, motivation, and a much-needed sense of camaraderie during this demanding period. In no particular order, I would like to thank \uline{\textbf{Goban}}, \uline{\textbf{Caroline}}, \uline{\textbf{Aryan}}, \uline{\textbf{Sarujan}}, \uline{\textbf{Ayesha}}, \uline{\textbf{Thomas}}, \uline{\textbf{Sahaj}}, \uline{\textbf{'Lavan}}, \uline{\textbf{Don}}, \uline{\textbf{Bethel}}, \uline{\textbf{Aman}}, \uline{\textbf{Sara}}, \uline{\textbf{Laura}}, \uline{\textbf{Adina}}, \uline{\textbf{Charlene}}, \uline{\textbf{Louisa}}, \uline{\textbf{Ceaser}}, \uline{\textbf{Obi}}, \uline{\textbf{Joelle}}, \uline{\textbf{Charlie}}, \uline{\textbf{Aqil}}, \uline{\textbf{Ben}}, \uline{\textbf{Karis}}, \uline{\textbf{Harry}}, \uline{\textbf{Thanawitch}}, \uline{\textbf{Margaret}}, \uline{\textbf{Xander}}, and \uline{\textbf{Adam}}. Your discussions, encouragement, and friendship have been truly appreciated.\newline

\noindent Love all of you truly so so much. Couldn't and wouldn't have made it here without any of you in my live. I give me deepest thanks.

\noindent I am also immensely grateful to my supervisor, \uline{\textbf{Sarah}}, for her invaluable guidance, insightful feedback, and consistent support throughout the duration of this dissertation. Her expertise and encouragement were instrumental in navigating the complexities of this research.\newline

\noindent Finally, I acknowledge the developers and maintainers of the open-source software and publicly available datasets that made this research possible.