% ai-phishing-detection-dissertation/report/sections/3-research-methodology/implementation-tools/user-interface-components.tex

\subsubsection*{User interface components}
The output of the model's predictions along with the relevant XAI explanations need to be rendered in a user-friendly format for post analysis, targeting users of all technical level. It takes inspiration from \cite{van2024applicability} and \cite{greco2023explaining} for UX design principles, to ensure that the explainability aspect of the system is informative, and most importantly, actionable.\newline

\noindent The Flask Web Framework is an important component for building a lightwight yet scalable web framework for Python. It is specifically suited for applications that do not require a full, cohesive web development template, and is classed as a microframework \citep{pallets2010}.

\begin{itemize}
  \item \textbf{Core architecture}:
  \begin{itemize}
    \item Utilises REST API endpoints for model predictions and XAI outputs.
    \item Uses JWT for authentication user access control with "\texttt{Flash-Security}".
  \end{itemize}
  \item \textbf{Phishing-specific site routes}:
  \begin{itemize}
    \item "\texttt{/analyse/email}": Takes the raw emails and returns user-friendly explanations.
    \item "\texttt{/analyse/url}": A visualisation of URL deception features via a heatmap or confusion matrix.
    \item "\texttt{/audit/log}": A log of decision histories to keep track for forensic uses.
  \end{itemize}
\end{itemize}

\noindent D3.js (alternatively known as D3), is an open-source JavaScript library specifically for interactive data visualisations through the use of SVG, HTML5, and CSS web components \citep{observable2025d3}.

\begin{itemize}
  \item \textbf{Interactive visualisations}:
  \begin{itemize}
    \item Use force-directed graphs for SHAP feature importance scores.
    \item Generate heatmaps for attention trends from DistilBERT.
    \item Create timeline views for time-based attacks.
  \end{itemize}
  \item \textbf{Usability optimisations}:
  \begin{itemize}
    \item Have a colour-blind friendly colour scheme based on the virdis scale \citep{sallan2021virdis}.
    \item Have a ELI5 toggle for very simplified views.
    \item DIsplay threat severity indicators that are clearly colour coded.
  \end{itemize}
\end{itemize}

\noindent Utilising another JavaScript based library, React, can help build intuitive and complex user interfaces with dynamic web components, which are modular in nature. It simplifies and allows for more efficient development with JSX syntax and a virtual DOM \citep{meta2025react}.

\begin{itemize}
  \item \textbf{Analyst dashboard}:
  \begin{itemize}
    \item Use a prioritisation queue that ranks cases by confidence scores -- from both model outcome and human participation.
    \item Have side-by-side comparisons of different XAI explanations.
    \item Have a reporting feature to flag false positives.
  \end{itemize}
  \item \textbf{Device-specific adaptations}:
  \begin{itemize}
    \item For different devices, have dynamic resizing like for SOC monitors or tablets.
    \item Have touch-optimised XAI interactive explanations, for further explorations.
  \end{itemize}
\end{itemize}
