\label{sec:Appendix 3}

\begin{longtable}{|p{5cm}|p{6cm}|p{4cm}|}
    \hline
    \textbf{Phase} & \textbf{Tasks} & \textbf{Duration} \\
    \hline
    \endfirsthead
    \hline
    \textbf{Phase} & \textbf{Tasks} & \textbf{Duration} \\
    \hline
    \endhead
    
    \hline
    \endfoot
    
    \hline
    \endlastfoot
    
    Research and Planning & 
    \begin{itemize}
        \item Conduct literature review.
        \item Refine research questions and objectives.
        \item Identify datasets (e.g., PhishTank, Kaggle).
    \end{itemize} & 
    End Jan – Early Feb (1–2 weeks) \\
    \hline
    
    Dataset Preparation & 
    \begin{itemize}
        \item Clean and preprocess data.
        \item Perform feature extraction (e.g., URLs, metadata).
        \item Split data into training, validation, and test sets.
    \end{itemize} & 
    Early Feb (1–2 weeks) \\
    \hline
    
    Model Development & 
    \begin{itemize}
        \item Develop baseline model (e.g., Random Forest, Logistic Regression).
        \item Experiment with advanced models (e.g., BERT).
        \item Train and validate models.
    \end{itemize} & 
    Mid-Feb – Early Mar (3 weeks) \\
    \hline
    
    Integration of XAI & 
    \begin{itemize}
        \item Implement SHAP and LIME for interpretability.
        \item Visualize explanations (e.g., feature importance, email components).
        \item Ensure explanation clarity and usability.
    \end{itemize} & 
    Early Mar – Mid-Mar (2 weeks) \\
    \hline
    
    Evaluation and Comparison & 
    \begin{itemize}
        \item Evaluate model performance (accuracy, F1-score).
        \item Assess interpretability using metrics (faithfulness, stability).
        \item Compare XAI-enhanced system to black-box models.
    \end{itemize} & 
    Mid-Mar – End Mar (2 weeks) \\
    \hline
    
    Report Writing & 
    \begin{itemize}
        \item Write methodology, results, and discussion sections.
        \item Proofread and finalize dissertation.
        \item Integrate appendices (e.g., Gantt chart, ethical approval).
    \end{itemize} & 
    April – Mid-May (5–6 weeks) \\
    \hline
    
    \end{longtable}
    